Gradual Soundness: Lessons from Static Python

### 2022-01-10 pj-7.1/paper.tex

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"

@@

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

@@

%\begin{figure}[t]
%  How to enforce type \code{List[int]} simply?
%  \begin{itemize}
%    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
%    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
%    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
%  \end{itemize}
%  2022-01-07: might be an interesting figure idea, showing the two options for enforcing a data structure
%    Q How to enforce data structure without traversing
%     \=> option 1: Shallow
%     \=> option 2: Deep
%  \caption{FILL illustrate list boundary and checks}
%  \label{f:list-boundary}
%\end{figure}

@@

@@

Cinder is a bytecode compiler that extends the Python instruction set.
All the standard Python instructions work similarly to Python 3.8;
in other words, an untyped Python program has the same semantics in
Cinder as in the standard interpreter.
The added instructions help either to express runtime checks or to
quickly perform a statically-determined action.

The two main instructions that express run-time checks are \code{CAST}
and \code{CHECK\_ARGS}.
The former checks a value against a type.
The latter is a special cast for functions and methods;
it appears at the top of every function and checks that the
actual parameters match the types of the formal parameters.

A simple optimizing instruction is \code{FAST\_LEN}, which expects
a built-in value\footnote{Because of subtyping, it does not suffice to check
the type. All Python lists support \code{FAST\_LEN} but some subclasses
of the Python list type may not.}
and quickly computes its length.
Another is \code{LOAD\_FIELD}, which scans a fixed offset within an object.
By contrast, the standard Python instruction for loading a field needs to
search a dictionary~(\code{LOAD\_ATTR}).

In addition to speeding code up, the \SP{} compiler takes care
to avoid slowing code down.
It does so by avoiding casts and run-time checks that are certain to succeed.
For example, calls from a typed context to a typed function are statically
checked for compatibility.
There is no need to run \code{CHECK\_ARGS} on typed inputs, so the bytecode
compiler inserts a special call instruction to skip past the check.

@@

@@

%% TODO 2022-01-04 (somewhere) explain that a verified compiler is an entirely separate project,
%%  need bytecode model, full transformations, memory model / machine etc.
%%  hence we don't model optimizations, little benefit
%%  opt. don't tell us about the surface language ... the particular SP opts are changing representations but not results

@@

To validate the design of \SP{}, we have developed a mechanized model using PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
    with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a tiny formal model
that illustrates the \SP{} gradual interaction strategy.
The second claim is justified by our Redex development;
in particular, by its test suite.
This section describes the Redex development in broad strokes.

\subsection{Goals and Non-Goals}

Our Redex model of \SP{} includes the key aspects of its types,
statements, and values.
The model has two main goals:
illustrate the boundaries where typed and dynamic code can mix,
and show how to soundly protect these boundaries.
Compared to other gradual languages, these boundaries are quite restrictive---which
partly explains how it is able to perform well in practice~(\cref{s:eval}).
Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021},
which is a compelling argument for their soundness.

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (e.g. primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}).
Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to the baseline Python semantics.
These features include \code{eval}, first-class classes, multiple inheritance, and module-level cells;
see \cref{s:impl} for further discussion.

Lastly, the model does not include type-directed optimizations.
Validating the correctness of optimizations is a separate task that requires
significant components; namely, a model of the \SP{} bytecode and a faithful
rendering of its type-based transformations.
Our model establishes soundness as a baseline.
Future work can use soundness to reason informally about optimizations,
or even take it as a starting point for compiler verification.

@@

%\begin{theorem}[Optimization Soundness]
%  If\ \(~\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
%  then\ \(~\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
%\end{theorem}

%\begin{theorem}[Optimization Effectiveness]
%  FILL how do we know optimizations are useful rather than trivial?
%  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
%  %%   for general heuristic opts.
%  %% - is [opt(e) != e] true for interesting programs?
%\end{theorem}

\subsection{Redex Development, Navigation, Integration, Conformance Tests}

%% - count tests
%% - summarize expressions
%% - 
%% - pbt?

\subsection{Bug Reports}

While formalizing Static Python, we submitted N bug reports to the language developers.
The developers acknowledged M of these bugs as issues to fix.

\begin{center}
  %% most filed by KC, a few by Ben
  %% + Carl https://github.com/facebookincubator/cinder/issues/64
  \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
\end{center}

@@

\subsection{Python Interaction}

@@

extend model with Python ... two languages ... boundaries ... argue for soundness

\subsection{Bytecode, Optimization}
\label{s:optimize}

@@

@@

@@

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%%   ... or at least mention both, be clear the focus is Dict b/c older
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"

@@

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

@@

%\begin{figure}[t]
%  How to enforce type \code{List[int]} simply?
%  \begin{itemize}
%    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
%    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
%    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
%  \end{itemize}
%  2022-01-07: might be an interesting figure idea, showing the two options for enforcing a data structure
%    Q How to enforce data structure without traversing
%     \=> option 1: Shallow; + works for legacy code; - doesn't guarantee full type
%     \=> option 2: Deep; + guarantees full type; - rejects legacy code
%     \=> (option 3: traverse/wrrap; + guarantees full type; + works for legacy code; - slow; - complex
%  \caption{FILL illustrate list boundary and checks}
%  \label{f:list-boundary}
%\end{figure}

@@

%% TODO need this para?
We present these in less detail than the type system highlights because
we suspect they are less-widely applicable than the type system features.
Some rely on an optimizing JIT compiler, for example.
Nevertheless, these highlights are an important part of what makes \SP{} successful.

@@

Cinder is a bytecode compiler that extends the Python instruction set.
All the standard Python instructions work similarly to Python 3.8;
in other words, an untyped Python program has the same semantics in
Cinder as in the standard interpreter.
The added instructions help either to express runtime checks or to
efficiently perform some action.
For example, the \code{FAST\_LEN} instruction quickly computes the
length of a built-in value.
%% TODO is it really undefined
\SP{} uses type information to decide where this instruction is appropriate
and applies it as an optimization.

@@

\subsubsection{Classloader}
\label{s:classloader}

\SP{} keeps track of typed functions and typed classes with a specialized classloader.
At runtime, the classloader keeps a registry of typed objects and whether or not they
are mutable.
The registry helps the bytecode reference objects and types via their module names.
The mutability bit lets \SP{} optimize access for immutable data while soundly allowing
patches to mutable data.
Overall, it is an essential piece of infrastructure to enable soundness and optimizations.

@@

@@

To validate the design of \SP{}, we have developed a mechanized model using PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
    with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a tiny formal model
that illustrates the \SP{} gradual interaction strategy.
%% Compared to other gradual languages, these boundaries are quite restrictive---which
%% partly explains how it is able to perform well in practice~(\cref{s:eval}).
%% Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021}.
The second claim is justified by our Redex development;
in particular, by its test suite.
This section describes the Redex development in broad strokes.

@@

\subsection{Redex Development, \SP{} Conformance}
%% pbt for soundness?

By contrast to the small model language presented in this section,
our Redex model covers both the typed--untyped boundaries in \SP{}
and a large subset of Python.
Notable Python features include module imports, loops, exception handlers, and
delete statements (for class members and attributes).
To ensure that these features are implemented correctly and soundly, the
model includes hundreds of test programs adapted from the \SP{} test suite.\footnote{
  Most of the N tests are direct translations of \SP{} conformance tests to
  the syntax of the model.
  A program that implements the translation is included with our Redex development.
  The other N tests are not directly expressible in the model; for these, we
  translated salient aspects of each test by hand.}
These tests exercise a variety of Python behaviors, thereby giving us confidence
that the model is sound for Python.
Some tests intentionally raise type errors, some raise exceptions, and the rest
compute a specific result.
We have additionally verified that the model and \SP{} agree on these end-to-end
behaviors.
Thus we believe that \SP{} is sound as well.
%% ... mention that SP is a large and hard-to-analyze language ... opaque?

The model intentionally does, however, not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (e.g. primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}).
Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to the baseline Python semantics.
These features include \code{eval}, first-class classes, multiple inheritance, and module-level cells;
see \cref{s:impl} for further discussion.
Lastly, the model does not include type-directed optimizations.
Validating the correctness of optimizations is a separate task that requires
significant components; namely, a model of the \SP{} bytecode and a faithful
rendering of its type-based transformations.
Our model establishes soundness as a baseline.
Future work can use soundness to reason informally about optimizations,
or even take it as a starting point for compiler verification.

\subsection{The Payoff: Issues Reported}

While modelling \SP{}, we found several issues with the language
and submitted them as bug reports to the \SP{} developers.
Overall, we reported 25 issues~(listed in \cref{a:github-issues}).
Four of these were critical soundness issues, one of which resulted in
a segmentation fault.
All but one of these soundness issue have been fixed.
Five issues were relatively minor; these dealt with confusing error messages
and incorrect tests.
The remaining 16 issues report bugs in language design and implementation.

In addition to these formal issue reports,
we had several long discussions together (between Brown and Instagram)
about finer points in the language design.

@@

\subsection{Interactions with Python}

@@

model with Python ... two languages ... boundaries ... argue for soundness

%% C-types are among the most complex because of casts? no not really complex, perhaps notable

\subsection{Bytecode and Optimizations}
\label{s:optimize}

\SP{} can generate more-efficent code than Python because it targets
the Cinder runtime.
Cinder extends Python with bytecode instructions to check types at run-time,
to construct \SP{} data structures, and to perform optimized actions.
\Cref{t:bytecode} lists a few representative bytecode instructions.

\begin{table}
  \caption{Cinder bytecode instructions}
  \label{t:bytecode}

  \begin{tabular}{lll}
    %% TODO remove _
    %% TODO pick instructions
    {Instruction} & {Purpose} & {Description} \\\hline
    \bcinst{CAST} & Soundness & Assert that a value matches a type \\
    \bcinst{CHECK\_ARGS} & Soundness & Cast all inputs to a function \\[1ex]

    \bcinst{BUILD\_CHECKED\_MAP} & Constructor & Make a checked dictionary \\
    %\bcinst{BUILD\_CHECKED\_LIST} & Constructor & Make a checked list \\
    \bcinst{TP\_ALLOC} & Constructor & Make a \SP{} object \\[1ex]

    \bcinst{INVOKE\_FUNCTION} & Optimization & Execute a direct function call \\
    \bcinst{INVOKE\_METHOD} & Optimization & Execute a method via the object's vtable \\
    \bcinst{LOAD\_FIELD} & Optimization & Read an object field \\
    \bcinst{STORE\_FIELD} & Optimization & Write to an object field \\
    \bcinst{FAST\_LEN} & Optimization & Get length of a built-in value \\
    \bcinst{REFINE\_TYPE} & Optimization & Type declaration for the JIT \\

%%% Other bytecodes ... either from Carl's talk or the SP repo
%CONVERT_PRIMITIVE
%INT_LOAD_CONST_OLD
%JUMP_IF_NONZERO_OR_POP
%JUMP_IF_ZERO_OR_POP
%LIST_DEL
%LOAD_ITERABLE_ARG
%LOAD_LOCAL
%LOAD_MAPPING_ARG
%POP_JUMP_IF_NONZERO
%POP_JUMP_IF_ZERO
%PRIMITIVE_BINARY_OP
%PRIMITIVE_BOX
%PRIMITIVE_COMPARE_OP
%PRIMITIVE_LOAD_CONST
%PRIMITIVE_UNARY_OP
%PRIMITIVE_UNBOX
%RETURN_PRIMITIVE
%SEQUENCE_GET
%SEQUENCE_REPEAT
%SEQUENCE_SET
%STORE_LOCAL

  \end{tabular}
\end{table}

The two instructions that express run-time checks are \code{CAST}
and \code{CHECK\_ARGS}.
The former checks a value against a type.
The latter is for functions and methods; it checks all inputs to a function
against their declared types.

By convention, a \code{CHECK\_ARGS} instruction always appears on the first
line of a typed function body.
The optimizer uses this convention to skip argument checks by jumping
past them when it is safe to do so; namely, whenever a typed function (or method)
gets called with well-typed arguments.
%% TODO ask C&D about materialize for fn calls

\Cref{t:bytecode} presents two instructions that allocate \SP{}-specific data structures.
The first creates a checked dictionary; the second creates an object with a vtable~(\cref{s:vtable}).
Cinder comes with a similar instruction to build checked lists.
The developers will add more instructions along the same lines in the future.

The remaining instructions are for optimization.
Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
to Python's dynamic call dispatch.
The former jumps to a statically-determined address;
the latter uses a v-table lookup to quickly find an address at run-time.
The load and store instructions for fields improve upon Python's generic attribute lookup.
In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
of using the Python \code{\_\_len\_\_} attribute.
Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
type is not clear from the context.

@@

%% TODO 2022-01-09 be careful here ... the model may handle module vars, but it's
%% just that they're untyped and (maybe) not tested

@@

JS++ is a sound gradual type system for JavaScript~\cite{jspp}.
By default, it separates JS++ from JavaScript values:
every value with a type must be defined in JS++ code,
and every value that may have come from Javascript has
a special type \code{external} to mark it as such.
Programmers can manually cast external values to a
more precise type.
The casts may fail at runtime, and may create a copy of
the original value if they succeed (depends on the type).
%% 2022-01-09 how does JS++ protect arrays sent to JS?

@@

\section{GitHub Issues}
\label{a:github-issues}

FILL if we're going to list issues, might as well report the titles, labels, and open/closed status.

The 25 GitHub issues that we filed during the formalization effort may be found at the following three links:

\begin{itemize}
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024} (N=20)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=4)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
\end{itemize}
